### 2021-12-10 manuscript.tex
Run quali results via Grammarly [ci skip]
This build is based on
https://github.com/lubianat/quali_phd/commit/cf629fe1f70f8f676f43e4ec04950c65a9674e6f.

This commit was created by the following CI build and job:
https://github.com/lubianat/quali_phd/commit/cf629fe1f70f8f676f43e4ec04950c65a9674e6f/checks
https://github.com/lubianat/quali_phd/actions/runs/1564915978

This manuscript
(\href{https://lubianat.github.io/quali_phd/v/8e4db1e93b09d5722f9c3655e461f238c65e0391/}{permalink})
was automatically generated
from \href{https://github.com/lubianat/quali_phd/tree/8e4db1e93b09d5722f9c3655e461f238c65e0391}{lubianat/quali\_phd@8e4db1e}
on December 8, 2021.

@@

The advent of single-cell technologies has ignited the desire of a deep knowledge on cells, the building blocks of life {[}\protect\hyperlink{ref-pNGap1Du}{1}{]}.
The Human Cell Atlas (HCA) project, has been a major player in the cell knowledge ecosystem, running since 2017 towards the task to characterize every cell type in the human body {[}\protect\hyperlink{ref-1GmbExweg}{2}{]}.
The HCA consortium gathers people from all over the world to tackle different parts of the project, so to have a diverse and equitable account of the cell type diversity. {[}\protect\hyperlink{ref-tjdjR2Xf}{3}{]}

Building a full atlas of human cells comes with multiple challenges. The project includes the detection, in single cells, of RNA species (scRNA-Seq), chromatin accessibility (scATAC-Seq), and protein markers (primarily by CYTOF), as well as spatial information on cells with multiplexed \emph{in situ} hybridization (such as MERFISH) and imaging mass cytometry {[}\protect\hyperlink{ref-1GmbExweg}{2},\protect\hyperlink{ref-kkwRTArg}{4}{]}. Every lab inside the project will contribute with its expertise, providing samples that are representative of human diversity.

HCA is set to revolutionize the biomedical sciences, by creating tools and standards for basic research, as well as allowing better characterization of disease, and thus, ultimately, improving diagnostics and therapy.
Its products (data, information, knowledge and wisdom) need to be FAIR: findable, accessible, interoperable and reusable.
Data stewardship and data management are growing as core dhttps://www.wikidata.org/wiki/Help:Multilingualemands of the scientific community, ranging from data management plans {[}\protect\hyperlink{ref-1DSEIjFha}{5}{]} to specialized personnel {[}\protect\hyperlink{ref-1DSEIjFha}{5}{]}.

The Human Cell Atlas has a dedicated team for organizing data: the Data Coordination Platform (DCP) {[}\protect\hyperlink{ref-zDRzmIGu}{6}{]} {[}\protect\hyperlink{ref-kkwRTArg}{4}{]}.
The DCP is responsible for tracing the plan for computational interoperability, from the data generators to the consumers.{[}\protect\hyperlink{ref-kkwRTArg}{4}{]}.
The Human Cell Atlas has its portal for data (\url{https://dahttps://www.wikidata.org/wiki/Help:Multilingualta.humancellatlas.org/}) which composes the data repository landscape with other resources, like the Broad Institute Single Cell Portal (\url{https://singlecell.broadinstitute.org/single_cell}) and the Chan-Zuckerberg Biohub Tabula Sapiens (\url{https://tabula-sapiens-portal.ds.czbiohub.org/}).
In addition to its core team, the HCA is poised to grow by community interaction, and states in its opening paper that ``As with the Human Genome Project, a robust plan will best emerge from wide-ranging scientific discussions and careful planning''.{[}\protect\hyperlink{ref-1GmbExweg}{2}{]}\\
Thus, this project inserts itself among the wide-ranging scientific discussions to improve data - and knowledge - interoperability.

The highlight of ``knowledge'' in the last paragraph is meant to stress that raw data \emph{per se} is not enough to turn the Atlas objectives into reality.
There is a long way from raw datasets to commonly agreed scientific knowledge.
And, ultimately, this long way is what allows humanity to take advantage of scientific endeavors.
Currently, the gap between data and knowledge is mostly targeted via the writing and sharing of scientific manuscripts, the \emph{de facto} currency of exchange of claims about the natural world.
The Human Cell Atlas Publication Commitee reviews and selects publications that are directly part of the HCA.
A set of publications is, thus, one of the major outputs of the whole endeavor.

The challenge that arises, thus, is one of managing a wealth of information and cast it into useful science.
Experimental articles that analyze thousands of cells pose an overload of information alone.
Ideally, we would like to understand, remember and make use of every statement produced by the HCA.
As this goal is humanely impossible, we need to develop tools to make the knowledge interoperable with the aid of computers.
At that point, the challenges of the HCA enter in resonance with the challenges of text-mining, biocuration and literature based discovery, which will be discussed in the chapter of this introduction.

@@

Given that a core goal of the Human Cell Atlas is to advance knowledge about \emph{all} human cell types, {[}\protect\hyperlink{ref-1GmbExweg}{2}{]} the definition of what a cell type is becomes important.
Although a number of views exist {[}\protect\hyperlink{ref-pNGap1Du}{1},\protect\hyperlink{ref-paKiKTRS}{7},\protect\hyperlink{ref-sWv3okaH}{8},\protect\hyperlink{ref-19yYJWEkz}{9},\protect\hyperlink{ref-BlxFp1v3}{10},\protect\hyperlink{ref-eOUB1Bmj}{11},\protect\hyperlink{ref-TupzTvtA}{12},\protect\hyperlink{ref-w0Woz7JE}{13},\protect\hyperlink{ref-srfuRMe}{14},\protect\hyperlink{ref-1GXQByEaQ}{15},\protect\hyperlink{ref-k0Pe9YOB}{16},\protect\hyperlink{ref-MoaYZAA0}{17},\protect\hyperlink{ref-WAeCfntu}{18}{]}, there is no formal, commonly agreed upon defintion of cell type.
A 2017 article on the Human Cell Atlas mentions{[}\protect\hyperlink{ref-BlxFp1v3}{10}{]}:

``Descriptors such as `cell type' and `cell state' can be difficult to define at the moment. An integrative, systematic effort by many teams of scientists working together and bringing different expertise to the problem could dramatically sharpen our terminology, and revolutionize the way we see our cells, tissues and organs. We invite you to join the effort.''
The article highlights both the current gap in knowledge and the need of a community effort to work in that direction, in a direction that justifies the existence of the present work.

One consequence of a lack of a definition is that there is no commonly agreed number of cell types, and not even on an order of magnitude.
As of November 2021, the leading answers in the Google Search Engine for the question ``How many different cell types are found in the human body?'' all point to around 200 different types (https://askabiologist.asu.edu/questions/human-cell-types, https://www.researchgate.net/post/How-many-cell-types-in-a-human-body-How-about-the-number-of-cell-cycles-in-each-species, https://www.kenhub.com/en/library/anatomy/types-of-cells-in-the-human-body), an estimate that is agreed upon by Bionumbers, a database of useful biological numbers {[}\protect\hyperlink{ref-SprCx2w6}{19}{]} (https://bionumbers.hms.harvard.edu/bionumber.aspx?id=103626).
A list of cell types in the adult human body on Wikipedia also amounts to around a couple hundred cell types {[}\protect\hyperlink{ref-uYuz0opI}{20}, =List\_of\_distinct\_cell\_types\_in\_the\_adult\_human\_body\&oldid=1044853788{]}.
However, the Cell Ontology has so far had catalogued 2,311 cell types of interest for the Human Cell Atlas as of June 2021 {[}\protect\hyperlink{ref-J5X2Xu6M}{21}{]}, increasing the estimate by at least one order of magnitude.
Additionally, with an estimate of 37 trillion cells on average per human body {[}\protect\hyperlink{ref-xWhvkd3B}{22}{]} and an ever-increasing report of new cell types/clusters in single-cell transcriptomics ({[}\protect\hyperlink{ref-HFiosSKK}{23}{]}), a precise estimate is not reasonable.
In fact, the Human Cell Atlas project itself does not commit to any estimates of numbers of cell types, due to the sheer difficulty of estimating a number given current knowledge. (Aviv Regev; reply to question in the HCA conference)

Even though there is no agreement, different views on cell types are maturing.
One core line of thought to define ``cell type'' is based on the cell type as an evolutionary unit defined by a Core Regulatory Complex (CoRC) of transcription factors. That definition enables the drawing of parallels, from the evolution of other biological entities (such as genes, proteins, and species) to cell types' evolution. Models of how multicellular life works greatly benefit from concepts such as ``sister types'' (cell types that diverged from a single ancestor), ``cell type homology'' (cell types in different species that share a common evolutionary origin), and ``cell type convergence'' (cell types that execute similar functions but which are not directly evolutionarily related) {[}\protect\hyperlink{ref-O95WN4x0}{24},\protect\hyperlink{ref-ogAGO9KH}{25}{]}

Another direction is based on the notion of attractors: regions of dynamical stability in a feature space, which might have different qualities. {[}\protect\hyperlink{ref-14J9lsoMd}{26},\protect\hyperlink{ref-OjxQDZpx}{27}{]} In this theory, ``basins of attraction'' direct cell phenotypes, providing points in, say, a gene expression space towards which different cells ``move'' their expression programs. This dynamic view see each cell type corresponding to ``a self-stabilizing regulatory program, which acts to maintain and restore the cell type-specific program of gene expression.'' {[}\protect\hyperlink{ref-19g5Iusf9}{28}{]} It alligns itself with dynamic systems theory, and some authors go as far as to say that ``Lacking the idea of attractors we have no clear idea of what a cell type is.'' {[}\protect\hyperlink{ref-4TJnL1xT}{29}{]}

As much as different concepts of species coexist {[}\protect\hyperlink{ref-g74dZjOM}{30}{]}, our quest to define cell types may take various forms. The challenge of representing cell types in the context of evolution is conceptually different from representing cell types in biomedical experimentation. In that second direction, the groundwork of the Cell Ontology {[}\protect\hyperlink{ref-1CI8uS6ka}{31},\protect\hyperlink{ref-gAysgnsS}{32},\protect\hyperlink{ref-17U3pGEjv}{33}{]} and CELDA {[}\protect\hyperlink{ref-RLHsA1U8}{34}{]} and the contributions of the International Workshop on Cells in Experimental Life Sciences series {[}\protect\hyperlink{ref-jX2ZmXcc}{35},\protect\hyperlink{ref-rK8KD0us}{36}{]} are notable.

Even though many sources of knowledge contribute to our knowledge about cell types {[}\protect\hyperlink{ref-16phbXNJE}{37}{]}, arguably single-cell transcriptomics is the workhorse for current efforts of the HUman cell Atlas, with an increasing amount of published studies using the methodology and of cells per study. {[}\protect\hyperlink{ref-16phbXNJE}{37}{]}
Current scRNA-seq data analyses often rely on unsupervised clustering of cells followed by assignment of cell-type labels to clusters.
For the clustering, bioinformaticians tailor parameter sets to a target resolution, i.e., the level of detail used to detect cell identities. {[}\protect\hyperlink{ref-WKbly37M}{38}{]} {[}\protect\hyperlink{ref-1GmbExweg}{2}{]}
When the clustering is finished, the groups of cells are annotated with class labels, representing the underlying biology in a language we can understand. {[}\protect\hyperlink{ref-ii1dVqoy}{39}{]}

Instead of assigning expression gates from pre-defined markers, as is the standard for flow-cytometry analysis, single-cell RNA-seq analysis pipelines usually start from \emph{de novo} clustering of cells followed by cluster annotation. {[}\protect\hyperlink{ref-WKbly37M}{38}{]}
While it is clear that clusters and cell types are different concepts {[}\protect\hyperlink{ref-WKbly37M}{38}{]}, often cluster labels are treated as cell types.
There are a number of ways to cluster cells to find groups of similarity, but arguably the current default is dirived on the methodology proposed by PhenoGraph. {[}\protect\hyperlink{ref-pEJYKNFc}{40}{]}
The protocol is to calculate the distances between cells in a reduced PCA space (with the number of dimensions chosen by the experimenters), followed by constructing a k-nearest-neighbours network, in which each cell is a node connected by \emph{k} (another parameter) edges to other cells.
Once the network is build, network modules (i.e.~cell clusters) are commonly found using the Louvain algorithm, published in 2008 by researchers of the Université catholique de Louvain, in Belgium. {[}\protect\hyperlink{ref-JpCsfe9G}{41}{]}
The cell clusters found by the PhenoGraph (or any other) algorithm are then labeled by domain experts, often based on genes differentially expressed on each cluster, so-called ``markers''. {[}\protect\hyperlink{ref-WKbly37M}{38}{]}

While it is possible to manually investigate the identities of which clusters, automatic methods have been developed to aid on the task. {[}\protect\hyperlink{ref-ii1dVqoy}{39}{]}
One approach (``marker-based automatic annotation'') bases itself on crossing clusters markers in the analyzed dataset with previous knowledge from databases like PanglaoDB {[}\protect\hyperlink{ref-1GezH4XNn}{42}{]} and CellMarker {[}\protect\hyperlink{ref-chGii6yw}{43}{]} {[}\protect\hyperlink{ref-ii1dVqoy}{39}{]}.
Another approach (reference-based automatic cell annotation) relies on base, expert-annotated datasets as references from which labels are transferred to the dataset of interest. {[}\protect\hyperlink{ref-ii1dVqoy}{39}{]}
Other methods bypass the clustering step and focus on labelling the individual cells, which avoids lumping dissimilar cells together, but require a high amount of reads per individual cells for it to be efficient. {[}\protect\hyperlink{ref-ii1dVqoy}{39}{]}
A recent review and tutorial by Clarke et al {[}\protect\hyperlink{ref-ii1dVqoy}{39}{]} provides an extensive account of current techniques.

Of note, even though a range of methods is available, the vast majority of techniques and publications do not use standard identifiers for cell types.
This is in contradiction with the acknowledgement by the community of the advantages of using identifiers the ad using standard identifiers, such as those provided by the Cell Ontology. {[}\protect\hyperlink{ref-ii1dVqoy}{39}{]} {[}\protect\hyperlink{ref-tvnuubVW}{44}{]} {[}\protect\hyperlink{ref-J5X2Xu6M}{21}{]} {[}\protect\hyperlink{ref-agnqfdk6}{45}{]} {[}\protect\hyperlink{ref-SozKx0lF}{46}{]} {[}\protect\hyperlink{ref-15xnotFsw}{47}{]}.
Nevertheless, projects that use Cell Ontology identifiers for single-cell RNA-seq data are appearing {[}\protect\hyperlink{ref-qT8WxqjA}{48}{]}, including python and R packages (e.g.~Besca {[}\protect\hyperlink{ref-P1gDLuog}{49}{]}, OnClass {[}\protect\hyperlink{ref-sW6aNZJB}{50}{]} and ontoProc{[}\protect\hyperlink{ref-15YmDXALp}{51}{]}), data management projects and reference datasets, (e.g.~Tabula Muris {[}\protect\hyperlink{ref-tdlWsvTd}{52}/{]} and Tabula Sapiens {[}\protect\hyperlink{ref-FjUqVIOp}{53}{]} Azimuth map {[}\protect\hyperlink{ref-T7RxnUNw}{54}/{]} and HubMap's ASCT+B Tables {[}\protect\hyperlink{ref-1FghWzORJ}{55}{]}) and annotation platforms (e.g.~the Cell Annotation Platform {[}\protect\hyperlink{ref-yMavNVkS}{56}{]} and CellTypist {[}\protect\hyperlink{ref-123DiCgc8}{57}{]}.

As elegantly put by Meehan et al {[}\protect\hyperlink{ref-UduMr7x2}{58}{]} the Cell Ontology is a ``manually constructed computer readable resource that links cell types by different relationships''.
it was first described in 2005 by Jonathan Bard, Seung Y Rhee† and Michael Ashburner {[}\protect\hyperlink{ref-1CI8uS6ka}{31}{]} and was oriented at creating an ``organism-independent classification of cells'', following criteria that included function, histology, lineage and ploidy and providing ``Cell-type unique identifiers (ID) that can be incorporated into any database holding cell-type-associated knowledge.'' It also had a didatic goal in itself, as the authors mention {[}\protect\hyperlink{ref-1CI8uS6ka}{31}{]}:
``It is designed to be useful in the sense that a researcher should be able to find, in a rapid and intuitive way, any cell type in any of the major model organisms and, having found it, learn a considerable amount about that cell type and its relationships to other biological objects.''
The collaborative project gradually evolved and changed its design and scope to fit new needs.
By 2011, for example, a need for computable definitions for hematopoietic cell types lead to a sizeable advance in the number and quality of immune cell types represented in CL. {[}\protect\hyperlink{ref-ecteO8vI}{59}{]}
It also included the addidion of species-specific cell types to better handle marker-based definitions, which are usually given at the species level. {[}\protect\hyperlink{ref-ecteO8vI}{59}{]}
Further developments over the years included both technical improvements as well of the addition of new cell types, and by the time of the last official CL publication, in 2016, it contained approximately 2,200 classes. {[}\protect\hyperlink{ref-agnqfdk6}{45}{]}\\
The Cell Ontology, currently, is growing as a resource for the Human Cell Atlas and in providing identifiers for cell types {[}\protect\hyperlink{ref-qT8WxqjA}{48}{]}.

In conclusion, the advancement of our \emph{formal} classification of cell types, such as in the Cell Ontology, represents a tangible goal of current cell-oriented large scale projects.
While purely theoretical developments have their value, refining the cell type theory in the context of knowledge management arguably will have a influence direcly on how the products of the Human Cell Atlas will impact modern science.
One reason is that formal systems enable automation of knowledge integration, and can feed intelligent systems that aid current research practices.
In the following chapter, it will be discussed how computer-based knowledge processing can influence life-sciences research, as well as discuss techniques and platforms to advance the frontier.

@@

The classification of biological concepts is at the core of biology. At least since the Aristotelian endeavours to group classes of animals, a good part of the scientific work is to capture concepts into knowledge systems {[}\protect\hyperlink{ref-DLTMbf54}{60}{]}.
Linnaeus' binomial system for naming species and Mendeleiev's periodic table are likely the two most famous classification systems, but are part of a much larger ecosystem of structuring scientific knowledge.

>>>>

This manuscript
(\href{https://lubianat.github.io/quali_phd/v/cf629fe1f70f8f676f43e4ec04950c65a9674e6f/}{permalink})
was automatically generated
from \href{https://github.com/lubianat/quali_phd/tree/cf629fe1f70f8f676f43e4ec04950c65a9674e6f}{lubianat/quali\_phd@cf629fe}
on December 10, 2021.

@@

The advent of single-cell technologies has ignited the desire for a deep knowledge of cells, the building blocks of life {[}\protect\hyperlink{ref-pNGap1Du}{1}{]}.
The Human Cell Atlas (HCA) project has been a major player in the cell knowledge ecosystem, running since 2017 to characterize every cell type in the human body {[}\protect\hyperlink{ref-1GmbExweg}{2}{]}.
The HCA consortium gathers people from all over the world to tackle different parts of the project to have a diverse and equitable account of the cell type diversity. {[}\protect\hyperlink{ref-tjdjR2Xf}{3}{]}

Building a complete atlas of human cells comes with multiple challenges. The project includes the detection, in single cells, of RNA species (scRNA-Seq), chromatin accessibility (scATAC-Seq), and protein markers (primarily by CYTOF), as well as spatial information on cells with multiplexed \emph{in situ} hybridization (such as MERFISH) and imaging mass cytometry {[}\protect\hyperlink{ref-1GmbExweg}{2},\protect\hyperlink{ref-kkwRTArg}{4}{]}. Every lab inside the project will contribute with its expertise, providing samples representing human diversity.

HCA is set to revolutionize the biomedical sciences by creating tools and standards for basic research, allowing better characterization of disease, and improving diagnostics and therapy.
Its products (data, information, knowledge and wisdom) need to be FAIR: findable, accessible, interoperable and reusable.
Data stewardship and management are growing as core demands of the scientific community, ranging from data management plans {[}\protect\hyperlink{ref-1DSEIjFha}{5}{]} to specialized data personnel {[}\protect\hyperlink{ref-1DSEIjFha}{5}{]}.

The Human Cell Atlas has a dedicated team for organizing data: the Data Coordination Platform (DCP) {[}\protect\hyperlink{ref-zDRzmIGu}{6}{]} {[}\protect\hyperlink{ref-kkwRTArg}{4}{]}.
The DCP is responsible for tracing the plan for computational interoperability, from the data generators to the consumers.{[}\protect\hyperlink{ref-kkwRTArg}{4}{]}.
The Human Cell Atlas has its portal for data {[}\protect\hyperlink{ref-kX6KnbUo}{7}{]}), which composes the data repository landscape with other resources, like the Broad Institute Single Cell Portal {[}https://singlecell.broadinstitute.org/single\_cell\textgreater) and the Chan-Zuckerberg Biohub Tabula Sapiens (\url{https://tabula-sapiens-portal.ds.czbiohub.org/}).
In addition to its core team, the HCA is poised to grow by community interaction. It states in its opening paper that ``As with the Human Genome Project, a robust plan will best emerge from wide-ranging scientific discussions and careful planning''.{[}\protect\hyperlink{ref-1GmbExweg}{2}{]}\\
Thus, this project inserts itself among the wide-ranging scientific discussions to improve data - and knowledge - interoperability.

The highlight of ``knowledge'' in the last paragraph is meant to stress that raw data \emph{per se} is not enough to turn the Atlas objectives into reality.
There is a long way from raw datasets to commonly-agreed scientific knowledge.
And, ultimately, this long way is what allows humanity to take advantage of scientific endeavours.
Currently, the gap between data and knowledge is mostly targeted by writing and sharing scientific manuscripts, the \emph{de facto} currency of exchange of claims about the natural world.
The Human Cell Atlas Publication Committee reviews and selects publications directly part of the HCA.
A set of publications is, thus, one of the significant outputs of the whole endeavour.

The challenge that arises, thus, is one of managing a wealth of information and casting it into useful science.
Experimental articles that analyze thousands of cells pose an overload of information alone.
Ideally, we would like to understand, remember, and use every statement produced by the HCA.
As this goal is humanely impossible, we need to develop tools to make the knowledge interoperable with the aid of computers.
At that point, the challenges of the HCA enter in resonance with the challenges of biocuration.
Knowledge representation tools will be discussed in the chapter on ontologies and knowledge graphs of this introduction.

@@

Given that a core goal of the Human Cell Atlas is to advance knowledge about \emph{all} human cell types, {[}\protect\hyperlink{ref-1GmbExweg}{2}{]} the definition of ``cell type'' becomes essential.
Although a number of views exist {[}\protect\hyperlink{ref-pNGap1Du}{1},\protect\hyperlink{ref-paKiKTRS}{8},\protect\hyperlink{ref-sWv3okaH}{9},\protect\hyperlink{ref-19yYJWEkz}{10},\protect\hyperlink{ref-BlxFp1v3}{11},\protect\hyperlink{ref-eOUB1Bmj}{12},\protect\hyperlink{ref-TupzTvtA}{13},\protect\hyperlink{ref-w0Woz7JE}{14},\protect\hyperlink{ref-srfuRMe}{15},\protect\hyperlink{ref-1GXQByEaQ}{16},\protect\hyperlink{ref-k0Pe9YOB}{17},\protect\hyperlink{ref-MoaYZAA0}{18},\protect\hyperlink{ref-WAeCfntu}{19}{]}, there is no formal, commonly agreed upon defintion of cell type.
A 2017 article on the Human Cell Atlas mentions{[}\protect\hyperlink{ref-BlxFp1v3}{11}{]}:

``Descriptors such as `cell type' and `cell state' can be difficult to define at the moment. An integrative, systematic effort by many teams of scientists working together and bringing different expertise to the problem could dramatically sharpen our terminology, and revolutionize the way we see our cells, tissues and organs. We invite you to join the effort.''

The article further highlights both the current gap in knowledge and the need for a community effort to work in finding definitions.

One consequence of a lack of a definition is that there is no commonly agreed number of cell types, and not even on the order of magnitude.
As of November 2021, the leading answers in the Google Search Engine for the question ``How many different cell types are found in the human body?'' all point to around 200 different types {[}\protect\hyperlink{ref-V85zvNGa}{20}, https://www.researchgate.net/post/How-many-cell-types-in-a-human-body-How-about-the-number-of-cell-cycles-in-each-species{]} {[}\protect\hyperlink{ref-WBDY0Uoy}{21}{]}, an estimate that is agreed upon by Bionumbers, a database of useful biological numbers {[}\protect\hyperlink{ref-SprCx2w6}{22}{]} {[}\protect\hyperlink{ref-jePJgR0H}{23}{]}.
A list of cell types in the adult human body on Wikipedia also amounts to around a couple hundred cell types {[}\protect\hyperlink{ref-uYuz0opI}{24}, =List\_of\_distinct\_cell\_types\_in\_the\_adult\_human\_body\&oldid=1044853788{]}.
However, the Cell Ontology has catalogued more than 2,311 cell types of interest for the Human Cell Atlas {[}\protect\hyperlink{ref-J5X2Xu6M}{25}{]}, increasing the estimate by at least one order of magnitude.
Additionally, with an estimate of 37 trillion cells on average per human body {[}\protect\hyperlink{ref-xWhvkd3B}{26}{]} and an ever-increasing report of new cell types/clusters in single-cell transcriptomics {[}\protect\hyperlink{ref-HFiosSKK}{27}{]}, it is reasonable that the number of relevant cell types is much larger.
The Human Cell Atlas project itself does not commit to any estimates of numbers of cell types due to the sheer difficulty of estimating a number given current knowledge. (Aviv Regev; reply to a question in the general HCA conference)

Even though there is no agreement, different views on cell types are maturing.
One core line of thought to define ``cell type'' is based on the cell type as an evolutionary unit.
That definition enables the drawing of parallels, from the evolution of other biological entities (such as genes, proteins, and species) to cell types' evolution.
Models of how multicellular life works greatly benefit from concepts as sister types (sharing a single ancestor), cell type homology ( sharing a common evolutionary origin), and cell type convergence (executing similar functions without direct evolutionary links) {[}\protect\hyperlink{ref-O95WN4x0}{28},\protect\hyperlink{ref-ogAGO9KH}{29}{]}

Another school of thought is based on attractors: regions of dynamical stability in a feature space. {[}\protect\hyperlink{ref-14J9lsoMd}{30},\protect\hyperlink{ref-OjxQDZpx}{31}{]}
In this theory, ``basins of attraction'' direct cell phenotypes, providing points in, say, a gene expression space towards which different cells ``move'' their expression programs.
This dynamic view sees each cell type corresponding to ``a self-stabilizing regulatory program, which acts to maintain and restore the cell type-specific program of gene expression.'' {[}\protect\hyperlink{ref-19g5Iusf9}{32}{]}
It aligns itself with dynamic systems theory, and some authors go as far as to say that ``Lacking the idea of attractors we have no clear idea of what a cell type is.'' {[}\protect\hyperlink{ref-4TJnL1xT}{33}{]}

As much as different species concepts coexist {[}\protect\hyperlink{ref-g74dZjOM}{34}{]}, the quest to define cell types may take various forms.
The challenge of representing cell types in evolution is conceptually different from representing cell types in biomedical experimentation.
In that second direction, the groundwork of the Cell Ontology {[}\protect\hyperlink{ref-1CI8uS6ka}{35},\protect\hyperlink{ref-gAysgnsS}{36},\protect\hyperlink{ref-17U3pGEjv}{37}{]} and CELDA {[}\protect\hyperlink{ref-RLHsA1U8}{38}{]} and the contributions of the International Workshop on Cells in Experimental Life Sciences series {[}\protect\hyperlink{ref-jX2ZmXcc}{39},\protect\hyperlink{ref-rK8KD0us}{40}{]} are notable.

\hypertarget{single-cell-transcriptomics}{%
\subsection{Single-cell transcriptomics}\label{single-cell-transcriptomics}}

Even though many sources of knowledge contribute to our understanding of cell types {[}\protect\hyperlink{ref-16phbXNJE}{41}{]}, arguably single-cell transcriptomics is the workhorse for current efforts of the HUman cell Atlas, with an increasing amount of published studies. {[}\protect\hyperlink{ref-16phbXNJE}{41}{]}
Current scRNA-seq data analyses often rely on unsupervised clustering of cells followed by the assignment of cell-type labels to clusters.
For the clustering, bioinformaticians tailor parameter sets to a target resolution, i.e., the level of detail used to detect cell identities. {[}\protect\hyperlink{ref-WKbly37M}{42}{]} {[}\protect\hyperlink{ref-1GmbExweg}{2}{]}
When the clustering is finished, the groups of cells are annotated with class labels, representing the underlying biology in a language we can understand. {[}\protect\hyperlink{ref-ii1dVqoy}{43}{]}

Instead of assigning expression gates from pre-defined markers, as is the standard for flow-cytometry analysis, single-cell RNA-seq analysis pipelines usually start from \emph{de novo} clustering of cells followed by cluster annotation. {[}\protect\hyperlink{ref-WKbly37M}{42}{]}
While it is clear that clusters and cell types are different concepts {[}\protect\hyperlink{ref-WKbly37M}{42}{]}, often cluster labels are treated as cell types.
There are several ways to cluster cells to find groups of similarity, but arguably the current default is derived from the methodology proposed by PhenoGraph. {[}\protect\hyperlink{ref-pEJYKNFc}{44}{]}
The protocol is to calculate the distances between cells in a reduced PCA space (with the number of dimensions chosen by the experimenters), followed by constructing a k-nearest-neighbours network.
Each cell is a node connected by \emph{k} (another parameter) edges to other cells.
Once the network is built, network modules (i.e.~cell clusters) are commonly found using the Louvain algorithm, published in 2008 by researchers of the Université Catholique de Louvain in Belgium. {[}\protect\hyperlink{ref-JpCsfe9G}{45}{]}
The cell clusters found by the PhenoGraph (or any other) algorithm are then labelled by domain experts, often based on genes differentially expressed on each cluster, so-called ``markers''. {[}\protect\hyperlink{ref-WKbly37M}{42}{]}

While it is possible to manually investigate the identities of which clusters, automatic methods have been developed to aid in the task. {[}\protect\hyperlink{ref-ii1dVqoy}{43}{]}
One approach (``marker-based automatic annotation'') bases itself on crossing clusters markers in the analyzed dataset with previous knowledge from databases like PanglaoDB {[}\protect\hyperlink{ref-1GezH4XNn}{46}{]} and CellMarker {[}\protect\hyperlink{ref-chGii6yw}{47}{]} {[}\protect\hyperlink{ref-ii1dVqoy}{43}{]}.
Another approach (reference-based automatic cell annotation) relies on base, expert-annotated datasets as references from which labels are transferred to the dataset of interest. {[}\protect\hyperlink{ref-ii1dVqoy}{43}{]}
Other methods bypass the clustering step and focus on labelling the individual cells, which avoids lumping different cells together but requires a high amount of reads per individual cells to be efficient. {[}\protect\hyperlink{ref-ii1dVqoy}{43}{]}
Clarke et al.'s recent review and tutorial {[}\protect\hyperlink{ref-ii1dVqoy}{43}{]} provides an extensive account of current techniques.

Of note, even though a range of methods is available, most techniques and publications do not use standard identifiers for cell types.
This is in contradiction with the acknowledgement by the community of the advantages of using identifiers the ad using standard identifiers, such as those provided by the Cell Ontology. {[}\protect\hyperlink{ref-ii1dVqoy}{43}{]} {[}\protect\hyperlink{ref-tvnuubVW}{48}{]} {[}\protect\hyperlink{ref-J5X2Xu6M}{25}{]} {[}\protect\hyperlink{ref-agnqfdk6}{49}{]} {[}\protect\hyperlink{ref-SozKx0lF}{50}{]} {[}\protect\hyperlink{ref-15xnotFsw}{51}{]}.
Nevertheless, projects that use Cell Ontology identifiers for single-cell RNA-seq data are appearing {[}\protect\hyperlink{ref-qT8WxqjA}{52}{]}, including python and R packages (e.g.~Besca {[}\protect\hyperlink{ref-P1gDLuog}{53}{]}, OnClass {[}\protect\hyperlink{ref-sW6aNZJB}{54}{]} and ontoProc{[}\protect\hyperlink{ref-15YmDXALp}{55}{]}), data management projects and reference datasets, (e.g.~Tabula Muris {[}\protect\hyperlink{ref-tdlWsvTd}{56}/{]} and Tabula Sapiens {[}\protect\hyperlink{ref-FjUqVIOp}{57}{]} Azimuth map {[}\protect\hyperlink{ref-T7RxnUNw}{58}/{]} and HubMap's ASCT+B Tables {[}\protect\hyperlink{ref-1FghWzORJ}{59}{]}) and annotation platforms (e.g.~the Cell Annotation Platform {[}\protect\hyperlink{ref-yMavNVkS}{60}{]} and CellTypist {[}\protect\hyperlink{ref-123DiCgc8}{61}{]}.

As elegantly put by Meehan et al. {[}\protect\hyperlink{ref-UduMr7x2}{62}{]} the Cell Ontology is a ``manually constructed computer-readable resource that links cell types by different relationships''.
it was first described in 2005 by Jonathan Bard, Seung Y Rhee† and Michael Ashburner {[}\protect\hyperlink{ref-1CI8uS6ka}{35}{]} and was oriented at creating an ``organism-independent classification of cells''.
It followed criteria that included function, histology, lineage and ploidy and providing ``Cell-type unique identifiers (ID) that can be incorporated into any database holding cell-type-associated knowledge.''
It also had a didactic goal in itself, as the authors mention {[}\protect\hyperlink{ref-1CI8uS6ka}{35}{]}:
``It is designed to be useful in the sense that a researcher should be able to find, in a rapid and intuitive way, any cell type in any of the major model organisms and, having found it, learn a considerable amount about that cell type and its relationships to other biological objects.''
The collaborative project gradually evolved and changed its design and scope to fit new needs.
By 2011, a need for computable definitions led to an advance in the number and quality of immune cell types represented in CL. {[}\protect\hyperlink{ref-ecteO8vI}{63}{]}
It also included the addition of species-specific cell types to handle better marker-based definitions, usually presented at the species level. {[}\protect\hyperlink{ref-ecteO8vI}{63}{]}
Further developments over the years included both technical improvements and the addition of new cell types.
By the time of the last official CL publication, in 2016, it contained approximately 2,200 classes. {[}\protect\hyperlink{ref-agnqfdk6}{49}{]}\\
Currently, the Cell Ontology is growing as a resource for the Human Cell Atlas and in providing identifiers for cell types {[}\protect\hyperlink{ref-qT8WxqjA}{52}{]} and presents over 2500 classes of cells (Figure \ref{fig:cl_counts})

\begin{figure}
\hypertarget{fig:cl_counts}{%
\centering
\includegraphics{https://github.com/lubianat/cell_ontology_count/blob/master/results/classes_per_release.png?raw=true}
\caption{Cell Ontology counts per release. Some releases around 2014 are not available anymore, thus yielding 0 counts.}\label{fig:cl_counts}
}
\end{figure}

While purely theoretical developments have their value, refining the cell type theory in the context of knowledge management arguably will directly influence how the products of the Human Cell Atlas will impact modern science.
One reason is that formal systems enable automation of knowledge integration and can feed intelligent systems that aid current research practices.
In the following chapter, it will be discussed how computer-based knowledge processing can influence life-sciences research and discuss techniques and platforms to advance the frontier.

@@

The classification of biological concepts is at the core of biology. At least since the Aristotelian endeavours to group classes of animals, a good part of the scientific work is to capture concepts into knowledge systems {[}\protect\hyperlink{ref-DLTMbf54}{64}{]}.
Linnaeus' binomial system for naming species and Mendeleev's periodic table are likely the two most famous classification systems but are part of a much larger ecosystem of structuring scientific knowledge.
